---
title: "Disorder_Per_Species"
output: html_document
date: "2024-11-15"
params:
  datadir: "/users/mirimia/izapata/projects/DisorderRegions/Obtain_DisDom/data"
  sumdir: "/users/mirimia/izapata/projects/DisorderRegions/data/Obtain_DisDom/Summary"
  DisAEX: "DisExA_AllSp.tab.gz"
  VastComp: "All-Disop_IupC_IupR_EX.tab.gz"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r libraries}
knitr::opts_chunk$set(echo = TRUE)
#Libraries
require(ggplot2) # figures
require("patchwork") # combine figures
require(tidyr)  # data table manipulation
require(dplyr)  # data table manipulation



# Params
# internal abbreviations of species
ordered_species = c("Hs2", "Mm2", "Bt2", "Mdo", "Gga", "Xtr", "Dre", "Cmi", "Bla", "Sp2", "Dme", "Eba", "Aae", "Bmo", "Tca", "Ame", "Bge", "Cdi", "Sma", "Obi")

# real abbreviations of species
real_species = c("Hs2"="Hsa", "Mm2"="Mmu", "Bt2"="Bta", "Mdo"="Mdo", "Gga"="Gga", "Xtr"="Xtr", "Dre"="Dre", "Cmi"="Cmi", "Bla"="Bla", "Sp2"="Spu", "Dme"="Dme", "Eba"="Eba", "Aae"="Aae", "Bmo"="Bmo", "Tca"="Tca", "Ame"="Ame", "Bge"="Bge", "Cdi"="Cdi", "Sma"="Sma", "Obi"="Obi")

# Specties in VastDB at present
vastdb_species = c("Hs2", "Mm2", "Bt2","Gga","Dre","Dme")

# Clades
Deuterostoma = c("Hs2", "Mm2", "Bt2", "Mdo", "Gga", "Xtr", "Dre", "Cmi", "Bla", "Sp2")
Protostoma = c("Dme", "Eba", "Aae", "Bmo", "Tca", "Ame", "Bge", "Cdi", "Sma", "Obi")

# Aesthetics

vastdv_colors=c("Hs2"="#00BFFF", "Mm2"="#159DFF", "Bt2"="#1873CD",
                "Gga"="#3F3F8B", "Dre"="#82359D","Dme"="red")


species_colors = c("Hs2"="#00BFFF", "Mm2"="#159DFF", "Bt2"="#1873CD", "Mdo"="#174B8B", "Gga"="#3F3F8B", "Xtr"="#64398B","Dre"="#82359D", "Cmi"="#9932CC", "Bla"="#2E8B57", "Sp2"="#96B12B", rev(c("Obi"="#FFFF00","Sma"="#FFCC33", "Cdi"="goldenrod", "Bge"="burlywood2", "Ame"="#FF9966", "Tca"="#FF6600","Bmo"="#CC6600", "Aae"="#993300", "Eba"="firebrick3", "Dme"="red")))

clade_colors = c("Vertebrata"= "mediumorchid", "Deuterostoma"="olivedrab3", 
                 "Insecta"="tan3", "Protostoma"="yellow", "Bilateria"="firebrick1", 
                 "Outgroups"="dimgray")
```


# Fig.3A

## Load data
Table with information on the disorder rate of the main exons (A) computed with:
- Disopred (and cutoff): "A_dDc", previously computed in the lab for VastDB
- Iupred2a (and cutoff): "A_dIc"
- Iupred2a (directly on raw values, not applying binarization step): "A_dIr"

Information of all AS exon skipping (EX) events (column "Event") from the 6 species (column "Species") currently present at VastDB that are also included in my analyses.


```{r Plots_build, fig.height=8}
# Load data
CompDisIup<-read.table(paste0(params$sumdir,"/",params$VastComp), 
                  stringsAsFactors = T, header=T, na.strings="na", sep = "\t",
                  quote = "",comment.char = "", fill = TRUE)
# Asign column names
colnames(CompDisIup)<-c("Species", "Event", "A_dDc","A_dIc","A_dIr")

# Data preparation
CompDisIup = CompDisIup %>%
  mutate(Species = factor(Species, levels=ordered_species))
```


## Plot
Figure 3A, density lines of the EX distribution rates for all species


Data formating: combine the 3 columns with information on the disorder rate into one, adding a column identifying with method corresponds with each value.

```{r}
# Still an improvement: using facet wrap
Comp2<- CompDisIup %>%
  rename(Disopred_cutoff = A_dDc,
         Iupred_cutoff = A_dIc,
         Iupred_raw = A_dIr) %>%
    pivot_longer(
    cols = c(Disopred_cutoff,Iupred_cutoff, Iupred_raw), 
    names_to = "Method",      
    values_to = "Disorder_Rate")
```


```{r}
# Plot
distributions<-ggplot(Comp2, aes(x = Disorder_Rate, color = Species, linetype = Species)) +
  geom_density(fill = NA) + 
  scale_color_manual(values = species_colors, labels=real_species) + 
  scale_linetype_manual(values=c("solid", "dashed", "dotted", "dotdash", "longdash", "twodash"),
                        labels=real_species) +
  labs(x ="disorder rate") +
  facet_wrap(~ Method) +
  theme_bw() +
  theme(legend.position = "none") 

# Print and save
distributions
##ggsave(paste0(params$sumdir,"/plots/Fig3a.pdf"),width=7,height=3)
```

# Fig3B

Pair-comparison of disorder scores computed with different methods (as density plot)



```{r}
#  Function to create denstity pots
graphXcomp2<-function(df, xname, yname){
    # parameters for the density plot
  binN<-30 # number of bins (density plot)
  lim1<-3  # limits for the colors: min (saturation)
  lim2<-5 # limits for the colors: min (saturation)
  
  # Methods to compare: transform name of the method to abbreviation (variable)
  ifelse(xname=="Iupred-cutoff",xcol<-df[,"A_dIc"],
         ifelse(xname=="Iupred-raw",xcol<-df[,"A_dIr"],
                ifelse(xname=="Disopred",xcol<-df[,"A_dDc"],"ERROR")))
  
  ifelse(yname=="Iupred-cutoff",ycol<-df[,"A_dIc"],
         ifelse(yname=="Iupred-raw",ycol<-df[,"A_dIr"],
                ifelse(yname=="Disopred",ycol<-df[,"A_dDc"],"ERROR")))
  

  # Graphs
  dens.plot <- ggplot(df, aes(x=xcol, y=ycol) ) +
    geom_hex(aes(fill= stat(log(count))), bins=binN) +
    geom_smooth(linewidth=0.75, linetype = 1, aes(col=Species)) +
    scale_color_manual(values=species_colors, labels=real_species) +
    scale_fill_viridis_c(limits=c(lim1,lim2), oob = scales::squish) + 
    labs(x=xname,y=yname) +
    # add tendency line
    geom_abline(intercept = 0, slope = 1, color="white",linetype=2,linewidth=0.5) +
    # one plot per species
    facet_wrap(~ Species, ncol=6, labeller = labeller(Species=real_species)) +
    theme_bw() +
    theme(legend.position = "none")
  
  return(dens.plot)
}

# Generate one plot for each pair-comparison (with all species per plot)  
all_Ir.Ic<-graphXcomp2(CompDisIup, xname="Iupred-raw",yname="Iupred-cutoff")
all_Ic.Dc<-graphXcomp2(CompDisIup, xname="Iupred-cutoff",yname="Disopred")
all_Ir.Dc<-graphXcomp2(CompDisIup, xname="Iupred-raw",yname="Disopred")

# Combine density lines (Fig. 3A) 
# with the 3 density plots of the pair-comparisons (Fig. 3B-D)
fig_all<- (distributions | all_Ic.Dc | all_Ir.Dc | all_Ir.Ic  ) +
  patchwork::plot_layout(nrow=4, guides="collect") & 
   patchwork::plot_annotation(tag_levels = "A") &
  theme(legend.position = "bottom", legend.direction = "horizontal",
        axis.text.x = element_text(angle = 0, hjust = 0.5, 
                                   vjust = 1, color="black",
                                   size=rel(0.85)),
        axis.text.y= element_text(color="black"))

# Print and save
fig_all
#ggsave(paste0(params$sumdir,"/plots/Fig3_ABCD.pdf"),width=10,height=9)

```

# Figure 4

Distribution (violin plot) of the disorder rates of all EX AS events for the 20 species, colored by clade. Computed with Iupred2a without cuttof (no probability binarization step): Iupred2a-raw).


## Load table 
Load table with information of the disorder rate for all 20 species.

```{r TypeEventSp}
# Load table
AdisT<-read.table(paste0(params$sumdir,"/",params$DisAEX), 
                  stringsAsFactors = T, header=F, na.strings="na", sep = "\t",
                  quote = "",comment.char = "", fill = TRUE)

# Asign column name
colnames(AdisT)<-c("Species","Gene","Event","A_Disr")

# Data preparation: species as ordered factors, add clade
AdisT = AdisT %>%
  mutate(Species = factor(Species, levels=ordered_species)) %>%
  mutate(Clade = ifelse(Species %in% Deuterostoma, "Deuterostoma", "Protostoma")) %>%
  mutate(Clade = factor(Clade, levels=c("Deuterostoma", "Protostoma")))
```

# Violin plot

```{r PlotFig4}
# Colored by clade
fig4<-ggplot(AdisT, aes(x=Species,y=A_Disr,fill=Clade, alpha=0.5)) + 
  geom_violin(trim=T) +
  geom_boxplot(width=0.25, show.legend = FALSE) +
  scale_fill_manual(values=clade_colors) +
  labs(x ="Species", y = "disorder score")+
  theme_bw() +
  theme(legend.position="right", legend.direction = "vertical",  
        axis.text.x = element_text(angle = 0, hjust = 0.5, vjust = 1, color="black"),
        axis.text.y = element_text(color="black")) +
  scale_x_discrete(labels=real_species) 

# Print and save
fig4 + guides(alpha=F)

#ggsave(paste0(params$sumdir,"/plots/Fig4.pdf"),width=8,height=3)

```


