---
title: "DomainPerSpecies"
output: html_document
date: "2024-11-18"
params:
  datadir: "/users/mirimia/izapata/projects/DisorderRegions/data"
  sumdir: "/users/mirimia/izapata/projects/DisorderRegions/data/Obtain_DisDom/Summary"
  DomA: "DomA_AllSp.tab.gz"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libraries}
knitr::opts_chunk$set(echo = TRUE)
#Libraries
require(ggplot2) # figures
require(tidyr)  # data manipulation
require(dplyr)  # data manipulation
require(stringr) # str_split, text manipulation in tables

# Params
# internal abbreviations of species
ordered_species = c("Hs2", "Mm2", "Bt2", "Mdo", "Gga", "Xtr", "Dre", "Cmi", "Bla", "Sp2", "Dme", "Eba", "Aae", "Bmo", "Tca", "Ame", "Bge", "Cdi", "Sma", "Obi")

# real abbreviations of species
real_species = c("Hs2"="Hsa", "Mm2"="Mmu", "Bt2"="Bta", "Mdo"="Mdo", "Gga"="Gga", "Xtr"="Xtr", "Dre"="Dre", "Cmi"="Cmi", "Bla"="Bla", "Sp2"="Spu", "Dme"="Dme", "Eba"="Eba", "Aae"="Aae", "Bmo"="Bmo", "Tca"="Tca", "Ame"="Ame", "Bge"="Bge", "Cdi"="Cdi", "Sma"="Sma", "Obi"="Obi")

# Specties in VastDB at present
vastdb_species = c("Hs2", "Mm2", "Bt2","Gga","Dre","Dme")

# Clades
Deuterostoma = c("Hs2", "Mm2", "Bt2", "Mdo", "Gga", "Xtr", "Dre", "Cmi", "Bla", "Sp2")
Protostoma = c("Dme", "Eba", "Aae", "Bmo", "Tca", "Ame", "Bge", "Cdi", "Sma", "Obi")

# Aesthetics

vastdv_colors=c("Hs2"="#00BFFF", "Mm2"="#159DFF", "Bt2"="#1873CD",
                "Gga"="#3F3F8B", "Dre"="#82359D","Dme"="red")


species_colors = c("Hs2"="#00BFFF", "Mm2"="#159DFF", "Bt2"="#1873CD", "Mdo"="#174B8B", "Gga"="#3F3F8B", "Xtr"="#64398B","Dre"="#82359D", "Cmi"="#9932CC", "Bla"="#2E8B57", "Sp2"="#96B12B", rev(c("Obi"="#FFFF00","Sma"="#FFCC33", "Cdi"="goldenrod", "Bge"="burlywood2", "Ame"="#FF9966", "Tca"="#FF6600","Bmo"="#CC6600", "Aae"="#993300", "Eba"="firebrick3", "Dme"="red")))

colTypes<-c("FE"="#E72A8A","PD"="#E6AB02","PU"="#A6761D","WD"="#7570B3") # overlap type
```


# Load table
Load table with informaiton of the functional domains (PFAM) detected to overlap with all exon skippin (EX) alternatively spliced (AS) events for the 20 bilaterian species.

```{r DomPerSp}
# Read File for human
DomSp<-read.table(paste0(params$sumdir,"/",params$DomA), 
                  stringsAsFactors = T, header=F, na.strings="na", sep = "\t",
                  quote = "",comment.char = "", fill = TRUE)

# Asign column names 
colnames(DomSp)<-c("Species","Event","Domain")

# Data manipulation:
# species as ordered factors, add clade, 
#add "detected" column, at least 1 overlap in the event? Yes/no

DomSp = DomSp %>%
  mutate(Species = factor(Species, levels=ordered_species)) %>%
  mutate(Clade = ifelse(Species %in% Deuterostoma, "Deuterostoma", "Protostoma")) %>%
  mutate(Clade = factor(Clade, levels=c("Deuterostoma", "Protostoma"))) %>%
  mutate(Detected = ifelse(Domain=="NO", "NO", "YES")) %>%
  mutate(Detected = factor(Detected, levels=c("NO","YES")))
```

# Fig 5A
Figure with stacked bars showing the number of events with ("YES") or withouth ("NO") at least one overlap with a functional domain.


```{r DomPerSp}
# Data preparation: Accumulation of events
YNdom<-DomSp %>%
  select(Species,Detected) %>%
  group_by(Species, Detected) %>%
  summarise(Freq = n()) %>%
  mutate(Species = factor(Species, levels=ordered_species)) %>%
  mutate(Clade = ifelse(Species %in% Deuterostoma, "Deuterostoma", "Protostoma")) %>%
  mutate(Clade = factor(Clade, levels=c("Deuterostoma", "Protostoma")))

# Compute the percentage of "YES" per species
percent<-YNdom %>%
  select(Species, Detected, Freq, Clade) %>%
  left_join(DomSp %>% 
              select(Species) %>%
              group_by(Species) %>%
              summarise(Total = n()), 
            by="Species") %>%
  mutate(Percentage=round(100*Freq/Total,2)) %>%
  filter(Detected=="YES") %>%
  select(Species, Percentage, Total, Clade)
```

Plot Fig. 5a

```{r}
# Plot: count, total number (text with "YES" proportions)
p5a<-ggplot(YNdom %>% rename(Domain=Detected)) +
  aes(x = Species, y = Freq, fill = Domain) +
  geom_bar(stat = "identity") + 
  geom_text(data=percent ,
            aes(x=Species, y=8000, label=round(Percentage,0), fill=NULL), 
            color="white", size=rel(2.5), angle=0) +
  labs(y="number") +
  scale_fill_manual(values=c("grey85","grey15")) +
  facet_wrap(~Clade, scales="free_x") +
  theme_bw() +
  theme(legend.position="none",  
        axis.text.x = element_text(angle = 0, hjust = 0.5, vjust = 1, color="black"),
        axis.text.y = element_text(color="black")) +
  scale_x_discrete(labels=real_species)
```

# Figure 5B: Overlaps

Data preparation: split the information of the domain in order to extract the name and type of overlap.

```{r NameDom}
# Getting overlap name and type (FE, WD, PU, PD)
df <- DomSp %>%
  select(Species,Event,Domain) %>%
  mutate(Domain = str_split(Domain, ",")) %>%
  unnest(Domain) %>%
  mutate(Name=str_extract(Domain, "=(.*?)=")) %>%
  mutate(Name=gsub("=","",Name)) %>%
  mutate(Type=str_extract(Domain, "=.*?=([A-Z]{2})")) %>%
  mutate(Type=gsub("=.*?=","",Type)) %>%
  mutate(Type = factor(Type)) %>%
  mutate(Name = factor(Name)) 

# Only consider events with at least one overlap,
# and add clade
dfType<-df %>%
  filter(Domain!="NO") %>%
  group_by(Species, Type) %>%
  summarise(Freq = n()) %>%
  mutate(Clade = ifelse(Species %in% Deuterostoma, "Deuterostoma", "Protostoma")) %>%
  mutate(Clade = factor(Clade, levels=c("Deuterostoma", "Protostoma")))
```

Plot Fig. 5B

```{r Plot_OverapeType}
# Plot proportions of each type of overlap, all species by Clade
p5b<-ggplot(dfType %>% rename(Overlap=Type)) +
  aes(x = Species, y = Freq, fill = Overlap) +
  geom_bar(stat = "identity") + 
  labs(y="number") +
  scale_fill_manual(values=c(colTypes)) +
  facet_wrap(~Clade, scales="free_x") +
  theme_bw() +
  theme(legend.position="none",  
        axis.text.x = element_text(angle = 0, hjust = 0.5, vjust = 1, color="black"),
        axis.text.y = element_text(color="black")) +
  scale_x_discrete(labels=real_species) 


# FE: the whole exon is contained in the domain
# WD: exon fully contains the domain
# PD: partial overlap of both, Downstream of event
# PU: partial overlap of both, Upstream of event
```

# Final figure 5

Combine both plots.

```{r}
# Combine A and B plots
fig5<-p5a/p5b + 
  patchwork::plot_layout(
    nrow=2
  ) +
  patchwork::plot_annotation(
    tag_levels = "A"
  ) &
  theme_bw() &
  theme(legend.position = "right", legend.direction = "vertical",
        axis.text.x = element_text(angle = 0, hjust = 0.5, vjust = 1, color="black"),
        axis.text.y = element_text(color="black"))  

# Print and save
fig5
#ggsave(paste0(params$sumdir,"/plots/Fig5.pdf"),width=8,height=5)
```



